---
title: "Case study: a real multi-donor, multi-cell-type cohort (GEO GSE60424)"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Case study: a real multi-donor, multi-cell-type cohort (GEO GSE60424)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(splitGraph)
```

The other vignettes use small synthetic tables. This one walks through a real
public cohort whose metadata has exactly the kind of structure that makes naive
random splits leak, and shows what splitGraph's validation, derivation, and
`split_spec` look like on it.

## The data

GEO series [GSE60424](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60424)
is an RNA-seq study of whole blood and six sorted immune cell populations from
20 donors: healthy controls and patients with type 1 diabetes, amyotrophic
lateral sclerosis, sepsis, or multiple sclerosis. The sample-level metadata
(not the expression data) ships with splitGraph as
`inst/extdata/GSE60424_samples.csv`; `inst/extdata/GSE60424_README.md` records
how it was derived from the GEO `characteristics` fields.

```{r load}
path <- system.file("extdata", "GSE60424_samples.csv", package = "splitGraph")
gse <- read.csv(path, stringsAsFactors = FALSE)
str(gse)
```

Three facts about its structure drive everything below:

```{r facts}
# 1. Every donor contributed several samples (one per cell population).
range(table(gse$subject_id))
# 2. Every donor was collected on its own date, so collection date (the natural
#    "batch") coincides with donor.
all(tapply(gse$batch_id, gse$subject_id, function(x) length(unique(x))) == 1)
# 3. Every donor has exactly one disease status.
all(tapply(gse$condition, gse$subject_id, function(x) length(unique(x))) == 1)
```

One caveat about the labels: GEO records multiple sclerosis samples as
"pretreatment" or "posttreatment", but these come from *different* donors
(three each), not from the same individuals over time. There is therefore no
repeated-measure time axis in this cohort, and we deliberately do **not** model
the labels as timepoints; they stay inside `disease_status`.

```{r no-longitudinal}
with(gse[gse$condition == "MS", ], table(subject_id, timepoint_id))
```

## Build the graph

Sorted cell populations are a categorical compartment of the sample, which is
what splitGraph's `Region` node type represents; disease status is the outcome.
`columns =` maps the CSV names onto the canonical ones.

```{r build}
meta <- gse[, c("sample_id", "subject_id", "batch_id", "cell_type", "condition", "sex")]
g <- graph_from_metadata(
  meta,
  columns = c(region_id = "cell_type", outcome_id = "condition"),
  graph_name = "GSE60424"
)
g
summary(g)$node_types
```

## What validation says

```{r validate}
report <- validate_graph(g)
report
summary(report)$by_code
```

Twenty `repeated_subject_samples` advisories: every donor is linked to several
samples, so any split that treats samples as independent will put the same
person on both sides. No error-severity issue, and no cross-study or cross-site
overlap, because each donor belongs to exactly one batch and one condition.

## How much would a naive split leak?

A plain random five-fold assignment of the 134 samples, without splitGraph:

```{r naive}
set.seed(1)
fold <- sample(rep(1:5, length.out = nrow(gse)))
straddling <- tapply(fold, gse$subject_id, function(f) length(unique(f)) > 1)
sum(straddling)
```

`r sum(straddling)` of 20 donors would appear in more than one fold. A model
evaluated that way sees each test donor's other cell populations during
training.

## Derive constraints

```{r derive}
by_subject <- derive_split_constraints(g, mode = "subject")
by_batch   <- derive_split_constraints(g, mode = "batch")
by_region  <- derive_split_constraints(g, mode = "region")
c(subject = by_subject$metadata$n_groups,
  batch   = by_batch$metadata$n_groups,
  region  = by_region$metadata$n_groups)
```

Because collection date and donor coincide, the batch partition *is* the
subject partition. Two groupings with different labels but the same partition
compare equal under a canonical relabelling:

```{r same-partition}
canon <- function(x) as.integer(match(x, unique(x)))
identical(canon(grouping_vector(by_subject)), canon(grouping_vector(by_batch)))
```

### Composite grouping and over-merging

Every donor spans every cell population, so a strict composite over subject
**and** region chains the whole cohort into one component: there is nothing
left to split.

```{r composite}
strict <- derive_split_constraints(g, mode = "composite", via = c("subject", "region"))
strict$metadata$n_groups
strict$metadata$warnings
```

`detect_dependency_components()` shows the component sizes directly, which is
the quickest way to see this coming before deriving:

```{r components}
comps <- detect_dependency_components(g, via = c("Subject", "Region"))
table(comps$table$component_size)
```

The rule-based strategy does not chain relations. With subject first in the
priority order, every sample has a subject, so region is never consulted and
the grouping is the subject partition, with region kept as an annotation:

```{r rule-based}
ruled <- derive_split_constraints(
  g, mode = "composite", strategy = "rule_based",
  via = c("subject", "region"), priority = c("subject", "region")
)
ruled$metadata$n_groups
head(as.data.frame(ruled)[, c("sample_id", "group_id", "constraint_type")], 3)
```

## The split_spec

Grouping by subject is the right primary constraint here; batch and region
travel as blocking annotations and the disease status as the stratum.

```{r spec}
spec <- as_split_spec(by_subject, graph = g)
spec
spec$block_vars
spec$stratum_var
head(as.data.frame(spec)[, c("sample_id", "group_id", "batch_group", "region_group", "stratum")], 7)
validate_split_spec(spec)
```

The leakage summary marks which of the validation findings this constraint
structurally severs:

```{r risks}
risks <- summarize_leakage_risks(g, constraint = by_subject, split_spec = spec)
unique(as.data.frame(risks)[, c("category", "severity", "severed")])
```

A consumer that groups on `group_id` and stratifies on `stratum` (for example
scikit-learn's `StratifiedGroupKFold` through the shipped Python reader, or
bioLeak's `as_leaksplits()`) will now keep every donor's cell populations on
one side of each fold while keeping the five conditions represented in every
fold as far as 20 donors allow.

## Working with a subset

Restricting the analysis to, say, healthy controls and multiple sclerosis
patients is a graph operation, not a re-import:

```{r subset}
keep <- gse$sample_id[gse$condition %in% c("Healthy Control", "MS")]
g_sub <- subset_graph(g, samples = keep, graph_name = "GSE60424: HC vs MS")
summary(g_sub)$node_types
spec_sub <- as_split_spec(derive_split_constraints(g_sub, "subject"), graph = g_sub)
table(spec_sub$sample_data$stratum)
```

## Handing off

```{r write, eval = requireNamespace("jsonlite", quietly = TRUE)}
out <- tempfile(fileext = ".json")
write_split_spec(spec, out)
validate_split_spec_json(out)
unlink(out)
```

The JSON carries the grouping, the blocking columns, the stratum, and the
provenance (`metadata$relations_used`, `splitgraph_version`, `derived_at`), so
the decision "one donor never straddles a fold" is recorded once and can be
executed anywhere.
