---
title: "Case study: a real multi-donor, multi-cell-type cohort (GEO GSE60424)"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Case study: a real multi-donor, multi-cell-type cohort (GEO GSE60424)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(splitGraph)
```

The other vignettes use small synthetic tables. This one walks through a real
public cohort whose metadata has exactly the kind of structure that makes naive
random splits leak, and shows what splitGraph's validation, derivation, and
`split_spec` look like on it.

## The data

GEO series [GSE60424](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60424)
is an RNA-seq study of whole blood and six sorted immune cell populations from
20 donors: healthy controls and patients with type 1 diabetes, amyotrophic
lateral sclerosis, sepsis, or multiple sclerosis. The sample-level metadata
(not the expression data) ships with splitGraph as
`inst/extdata/GSE60424_samples.csv`; `inst/extdata/GSE60424_README.md` records
how it was derived from the GEO `characteristics` fields.

```{r load}
path <- system.file("extdata", "GSE60424_samples.csv", package = "splitGraph")
gse <- read.csv(path, stringsAsFactors = FALSE)
str(gse)
```

Three facts about its structure drive everything below:

```{r facts}
# 1. Every donor contributed six or seven samples, one per cell population
#    that was successfully sorted for them.
range(table(gse$subject_id))
table(table(gse$subject_id))
# 2. Every donor was collected on its own date, so collection date (the natural
#    "batch") coincides with donor.
all(tapply(gse$batch_id, gse$subject_id, function(x) length(unique(x))) == 1)
# 3. Every donor has exactly one disease status.
all(tapply(gse$condition, gse$subject_id, function(x) length(unique(x))) == 1)
```

One caveat about the labels: GEO records multiple sclerosis samples as
"pretreatment" or "posttreatment", but these come from *different* donors
(three each), not from the same individuals over time. There is therefore no
repeated-measure time axis in this cohort, and we deliberately do **not** model
the labels as timepoints; they stay inside `disease_status`.

```{r no-longitudinal}
with(gse[gse$condition == "MS", ], table(subject_id, timepoint_id))
```

## Build the graph

Sorted cell populations are a categorical compartment of the sample, which is
what splitGraph's `Region` node type represents; disease status is the outcome.
`columns =` maps the CSV names onto the canonical ones.

```{r build}
meta <- gse[, c("sample_id", "subject_id", "batch_id", "cell_type", "condition", "sex")]
g <- graph_from_metadata(
  meta,
  columns = c(region_id = "cell_type", outcome_id = "condition"),
  graph_name = "GSE60424"
)
g
summary(g)$node_types
```

At 186 nodes this is past the size where drawing the whole graph tells you
anything. `focus = "ego"` is the view for a graph like this: it zooms to one
node's neighbourhood, so you can check a single donor's structure instead of
squinting at the cohort. Two hops out from donor `D20` reach its samples, and
through them the cell populations, collection date and disease status those
samples carry:

```{r ego, fig.width = 7.2, fig.height = 5, dpi = 150, out.width = "100%"}
plot(g, focus = "ego", node = "subject:D20", order = 2,
     legend_position = "bottomleft")
```

## What validation says

```{r validate}
report <- validate_graph(g)
report
summary(report)$by_code
```

Twenty `repeated_subject_samples` advisories, one per donor: every donor is
linked to several samples, so any split that treats samples as independent will
put the same person on both sides. Nothing rises above advisory, and the report
is short for two different reasons worth separating.

The cross-study and cross-site rules cannot fire because this graph has no
`Study` or `Site` nodes at all — the CSV carries neither, so those axes simply do
not exist here. `heavy_batch_reuse` does not fire because no batch holds half
the cohort: each collection date covers one donor's six or seven samples.

There is also no rule for a donor spanning several *regions*, and in this cohort
every donor spans six or seven of them. That is deliberate rather than an
oversight: one person contributing several sorted cell populations is the design
of the experiment, not an anomaly. The advisory that a donor has several samples
already carries the leakage signal; which compartments those samples came from
is information for the split, not a finding against the data.

## How much would a naive split leak?

A plain random five-fold assignment of the 134 samples, without splitGraph:

```{r naive}
set.seed(1)
fold <- sample(rep(1:5, length.out = nrow(gse)))
straddling <- tapply(fold, gse$subject_id, function(f) length(unique(f)) > 1)
sum(straddling)
```

`r sum(straddling)` of 20 donors would appear in more than one fold. A model
evaluated that way sees each test donor's other cell populations during
training.

## Derive constraints

```{r derive}
by_subject <- derive_split_constraints(g, mode = "subject")
by_batch   <- derive_split_constraints(g, mode = "batch")
by_region  <- derive_split_constraints(g, mode = "region")
c(subject = by_subject$metadata$n_groups,
  batch   = by_batch$metadata$n_groups,
  region  = by_region$metadata$n_groups)
```

Because collection date and donor coincide, the batch partition *is* the
subject partition. Two groupings with different labels but the same partition
compare equal under a canonical relabelling:

```{r same-partition}
canon <- function(x) as.integer(match(x, unique(x)))
identical(canon(grouping_vector(by_subject)), canon(grouping_vector(by_batch)))
```

The region partition is a different animal, and it is worth seeing why it is the
wrong split unit here even though it is a perfectly valid one. Grouping by cell
population cuts the cohort *across* donors rather than between them, so every
donor lands in six or seven different groups — the exact leak the study is
exposed to:

```{r region-wrong}
region_groups <- grouping_vector(by_region)
donor_spread <- tapply(region_groups[gse$sample_id], gse$subject_id,
                       function(x) length(unique(x)))
range(donor_spread)
sum(donor_spread > 1)   # donors split across more than one region group
```

Which is why region travels on the spec as a blocking annotation below, not as
`group_var`. The same column can be the right answer or the wrong one depending
on what has to generalise; the graph makes the difference inspectable rather
than a matter of taste.

### Composite grouping and over-merging

Six of the seven cell populations were sorted for all twenty donors. That is
enough for a strict composite over subject **and** region to chain the entire
cohort into one component — donor A's B-cells link to every other donor's
B-cells, which link to their own other populations, and so on until nothing is
left to split.

```{r composite}
strict <- derive_split_constraints(g, mode = "composite", via = c("subject", "region"))
strict$metadata$n_groups
strict$metadata$warnings
```

The empty warning vector is the part to notice. One group covering all 134
samples is a correct answer to the question that was asked, and `splitGraph`
does not flag it, so on a composite derivation read `metadata$n_groups` yourself
before going further. `detect_dependency_components()` shows it coming before
you derive anything:

```{r components}
comps <- detect_dependency_components(g, via = c("Subject", "Region"))
table(as.data.frame(comps)$component_size)
```

A single component of 134 — the cohort has no internal boundary along those two
relations together.

The rule-based strategy does not chain relations. With subject first in the
priority order, every sample has a subject, so region is never consulted and
the grouping is the subject partition, with region kept as an annotation:

```{r rule-based}
ruled <- derive_split_constraints(
  g, mode = "composite", strategy = "rule_based",
  via = c("subject", "region"), priority = c("subject", "region")
)
ruled$metadata$n_groups
head(as.data.frame(ruled)[, c("sample_id", "group_id", "constraint_type")], 3)
```

## The split_spec

Grouping by subject is the right primary constraint here; batch and region
travel as blocking annotations and the disease status as the stratum.

```{r spec}
spec <- as_split_spec(by_subject, graph = g)
spec
spec$block_vars
spec$stratum_var
head(as.data.frame(spec)[, c("sample_id", "group_id", "batch_group", "region_group", "stratum")], 7)
validate_split_spec(spec)
```

The leakage summary marks which of the validation findings this constraint
structurally severs:

```{r risks}
risks <- summarize_leakage_risks(g, constraint = by_subject, split_spec = spec)
unique(as.data.frame(risks)[, c("category", "severity", "severed")])
```

A consumer that groups on `group_id` and stratifies on `stratum` (for example
scikit-learn's `StratifiedGroupKFold` through the shipped Python reader, or
bioLeak's `as_leaksplits()`) will now keep every donor's cell populations on
one side of each fold while keeping the five conditions represented in every
fold as far as 20 donors allow.

## Working with a subset

Restricting the analysis to, say, healthy controls and multiple sclerosis
patients is a graph operation, not a re-import:

```{r subset}
keep <- gse$sample_id[gse$condition %in% c("Healthy Control", "MS")]
g_sub <- subset_graph(g, samples = keep, graph_name = "GSE60424: HC vs MS")
summary(g_sub)$node_types
spec_sub <- as_split_spec(derive_split_constraints(g_sub, "subject"), graph = g_sub)
table(spec_sub$sample_data$stratum)
```

## Handing off

```{r write, eval = requireNamespace("jsonlite", quietly = TRUE)}
out <- tempfile(fileext = ".json")
write_split_spec(spec, out)
validate_split_spec_json(out)$valid
unlink(out)
```

The JSON carries the grouping, the blocking columns, the stratum, and the
provenance (`metadata$relations_used`, `splitgraph_version`, `derived_at`), so
the decision "one donor never straddles a fold" is recorded once and can be
executed anywhere.
