---
title: "FAQ and design notes"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{FAQ and design notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(splitGraph)
```

## Why not just call `make_split_plan(group = , batch = )` directly?

For a dataset with one subject column and one batch column you should: bioLeak's
`make_split_plan()` reaches the same grouping in one call, and splitGraph adds a
hop. splitGraph earns its place when the structure is not one clean column per
axis:

- **Validation before splitting.** A sample linked to two subjects, a subject
  that appears in two studies or at two sites, a `time_index` that contradicts
  `timepoint_precedes`: `validate_graph()` reports these before any fold exists.
  Column-based grouping silently produces folds from inconsistent metadata.
- **Composite semantics.** `mode = "composite"` merges samples into connected
  components over *all* chosen relations. bioLeak's `combined` mode instead
  picks a primary axis and deletes training rows sharing a secondary level with
  the test set. Same metadata, different partitions; splitGraph makes the choice
  explicit and reproducible.
- **Pairwise relations.** Genetic relatedness and spatial adjacency are graded
  properties of *pairs*, not categories. `relatedness_edges_from_kinship()` and
  `spatial_edges_from_coords()` threshold them into edges and the derivation
  takes connected components. No categorical column can express this.
- **Provenance and interchange.** The `split_spec` records which relations,
  thresholds and strategy produced each group, and travels as schema-checked
  JSON to Python or any other consumer.

## When does composite-strict over-merge?

Strict composite grouping is transitive closure. If sample A shares a subject
with B, and B shares a batch with C, then A, B and C land in one group even
though A and C share nothing directly. With a few large batches this collapses
most of the dataset into one component:

```{r overmerge}
meta <- data.frame(
  sample_id  = paste0("S", 1:6),
  subject_id = c("P1", "P1", "P2", "P2", "P3", "P3"),
  batch_id   = c("B1", "B2", "B2", "B3", "B3", "B1"),
  stringsAsFactors = FALSE
)
g <- graph_from_metadata(meta)
table(grouping_vector(derive_split_constraints(g, "composite", via = c("subject", "batch"))))
```

Every subject bridges two batches, so all six samples form a single group and no
split is possible. Three remedies, in order of preference:

1. Ask whether every relation really *must* be severed. Grouping by subject
   alone here yields three groups; batch can be handled as a blocking
   annotation instead (`spec$block_vars`).
2. Use `strategy = "rule_based"`: each sample is grouped by the first relation
   in `priority` that is available to it, so relations do not chain.
3. Use `detect_dependency_components()` to see the component sizes before
   deriving, and `summarize_leakage_risks()` to see which leakage paths a given
   mode actually severs (`severed` column).

```{r remedies}
grouping_vector(derive_split_constraints(g, "composite", strategy = "rule_based",
                                         via = c("subject", "batch"),
                                         priority = c("subject", "batch")))
```

## How do thresholds interact with transitive closure?

`relatedness_edges_from_kinship(pairs, threshold)` keeps a pair when its kinship
is *at least* the threshold; `spatial_edges_from_coords(coords, radius)` keeps a
pair when its distance is *at most* the radius. The derivation then forms
connected components over the kept edges. Two consequences:

- Lowering a kinship threshold (or raising a radius) can only merge groups,
  never split them, and the merging is not local: a chain of individually
  just-over-threshold pairs joins its endpoints even if they are unrelated.
- The threshold is a property of the *edge set*, recorded on it and carried
  into `graph$metadata$edge_sources` and `spec$metadata$threshold`, so a
  reader of the spec can see which cut produced the groups.

```{r threshold}
pairs <- data.frame(id1 = c("P1", "P2"), id2 = c("P2", "P3"), kinship = c(0.26, 0.13))
meta <- data.frame(sample_id = c("S1", "S2", "S3"), subject_id = c("P1", "P2", "P3"))
build <- function(threshold) {
  g <- build_dependency_graph(
    list(create_nodes(meta, "Sample", "sample_id"), create_nodes(meta, "Subject", "subject_id")),
    list(create_edges(meta, "sample_id", "subject_id", "Sample", "Subject", "sample_belongs_to_subject"),
         relatedness_edges_from_kinship(pairs, threshold = threshold))
  )
  grouping_vector(derive_split_constraints(g, "relatedness"))
}
build(0.25)  # only P1~P2 pass: {S1,S2}, {S3}
build(0.10)  # P2~P3 also passes and chains: {S1,S2,S3}
```

## What does the `stratum` column mean, and does splitGraph stratify?

No. `stratum` is an *annotation*: the outcome level attached to each sample
(from `sample_has_outcome`, or the subject's outcome via `subject_has_outcome`).
It is exposed through `spec$stratum_var` so a consumer such as scikit-learn's
`StratifiedGroupKFold` or bioLeak's `stratify = TRUE` can balance folds.
Balancing is execution and belongs downstream; splitGraph only describes.

## Schema versioning policy, in one place

- `schema_version` (currently `r splitGraph:::.depgraph_schema_version`) is
  independent of the package version. It describes the on-disk JSON contract.
- The **major** component is the compatibility boundary. Files sharing the
  installed major load silently; readers fill fields missing from older files
  with `NA` and ignore unknown ones. A differing major loads with a warning
  suggesting `migrate_dependency_graph_json()` / `migrate_split_spec_json()`.
- Additive changes (a new column, a new metadata field) bump the minor
  component. Only a rename or a change of meaning would bump the major, and
  none has happened.
- The formal JSON Schemas live under `inst/schema/<schema_version>/`, and every
  written file carries a `$schema` URL pointing at its own version, so the
  reference stays valid after later bumps. `validate_graph_json()` and
  `validate_split_spec_json()` (or `read_*(validate = TRUE)`) check a file
  against the installed schema without a JSON Schema engine.

## Can I build a graph straight from a SummarizedExperiment?

Yes. `graph_from_metadata()` is an S3 generic, and the
`SummarizedExperiment` method reads `colData()` as the metadata table. When
`colData` has no `sample_id` column the assay column names are used, so a
typical Bioconductor object needs no preparation at all. Pass
`sample_id_col =` to name a different column, and `columns =` to map your own
names onto the canonical ones exactly as for a data frame.

```{r se, eval = requireNamespace("SummarizedExperiment", quietly = TRUE)}
meta <- data.frame(
  sample_id  = c("S1", "S2", "S3", "S4"),
  subject_id = c("P1", "P1", "P2", "P2"),
  batch_id   = c("B1", "B2", "B1", "B2"),
  stringsAsFactors = FALSE
)
se <- SummarizedExperiment::SummarizedExperiment(
  assays  = list(counts = matrix(0, nrow = 3, ncol = 4,
                                 dimnames = list(NULL, meta$sample_id))),
  colData = meta[, c("subject_id", "batch_id")]
)

g_se <- graph_from_metadata(se, graph_name = "from-se")
grouping_vector(derive_split_constraints(g_se, "subject"))

# identical to building from the data frame directly
identical(
  grouping_vector(derive_split_constraints(g_se, "subject")),
  grouping_vector(derive_split_constraints(graph_from_metadata(meta), "subject"))
)
```

## Which errors can I catch programmatically?

Every error inherits from `splitgraph_error` and carries a `code`; see
`?splitgraph_conditions` for the subclasses (`splitgraph_schema_error`,
`splitgraph_reference_error`, `splitgraph_ambiguity_error`,
`splitgraph_validation_error`, `splitgraph_io_error`).

```{r conditions}
g <- graph_from_metadata(data.frame(sample_id = c("S1", "S2"), subject_id = c("P1", "P2")))
tryCatch(
  query_neighbors(g, "sample:S9"),
  splitgraph_reference_error = function(e) e$code
)
```

## How large a cohort can splitGraph handle?

Every step is linear in nodes plus edges. The shipped benchmark
(`inst/bench/pipeline.R`) builds, validates, derives a default composite
constraint, enriches a spec and writes both JSON files for 20,000 samples in
seconds on a laptop; `tests/testthat/test-performance.R` fails if any step
regresses to superlinear behaviour. The one inherently quadratic output is the
explicit sample-pair table of `detect_shared_dependencies()` and
`detect_dependency_components()$metadata$projection_edges`, whose size is the
number of pairs sharing a target; grouping itself never enumerates pairs.
