---
title: "Modeling site, platform, relatedness, and spatial structure"
author: "Selçuk Korkmaz"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Modeling site, platform, relatedness, and spatial structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(splitGraph)
```

Beyond the classic subject / batch / study / time relations, `splitGraph`
models several further leakage axes, in two families:

- **Cluster-style relations** — collection **site**, tissue / anatomical
  **region**, sequencing **platform**, and **assay** — are categorical
  groupings. Each is auto-detected from a metadata column and handled by its own
  constraint mode, exactly like `subject` or `batch`.
- **Pairwise relations** — genetic **relatedness** and **spatial** proximity —
  are continuous and defined *between pairs*. They are modelled as thresholded
  edges and grouped by transitive closure, a partition a single categorical
  column cannot express.

This vignette builds and groups by each, and shows how the threshold drives the
pairwise grouping.

# Cluster-style relations: site, region, platform, assay

`graph_from_metadata()` auto-detects `site_id`, `region_id`, `platform_id`, and
`assay_id` columns and builds the corresponding typed nodes and edges. Each then
has its own constraint mode. The example below uses site, platform, and assay;
`region` behaves identically (a `region_id` column and `mode = "region"`) and is
omitted only to keep the output short.

```{r cluster}
meta <- data.frame(
  sample_id   = paste0("S", 1:6),
  subject_id  = c("P1", "P1", "P2", "P2", "P3", "P3"),
  site_id     = c("NYC", "NYC", "BOS", "BOS", "NYC", "BOS"),
  platform_id = c("illumina", "illumina", "nanopore", "nanopore", "illumina", "nanopore"),
  assay_id    = c("rnaseq", "rnaseq", "rnaseq", "wgs", "wgs", "wgs"),
  stringsAsFactors = FALSE
)

g <- graph_from_metadata(meta, graph_name = "structure-demo")

grouping_vector(derive_split_constraints(g, mode = "site"))
grouping_vector(derive_split_constraints(g, mode = "platform"))
grouping_vector(derive_split_constraints(g, mode = "assay"))
```

Site structure also has a validation rule of its own. A subject whose samples
were collected at more than one site is flagged, because grouping by site alone
would then place one individual on both sides of a split:

```{r cross-site}
report <- validate_graph(g)
report$issues[report$issues$code == "subject_cross_site_overlap",
              c("severity", "message")]

# Which constraint modes actually sever that path:
risks <- summarize_leakage_risks(g, constraint = derive_split_constraints(g, "subject"))
as.data.frame(risks)[as.data.frame(risks)$category == "subject_cross_site_overlap",
                     c("category", "severed")]
```

Whatever mode is primary, every detected cluster relation is also carried into
the `split_spec` as a *blocking annotation*, so a downstream consumer can block
on site, platform, or assay even when the split unit is something else — here,
subject:

```{r block-annotations}
spec <- as_split_spec(derive_split_constraints(g, mode = "subject"), graph = g)
spec$block_vars
head(spec$sample_data[, c("sample_id", "group_id",
                          "site_group", "platform_group", "assay_group")])
```

Any of these relations can also participate in a **composite** derivation, where
several dependency sources are combined and each connected component becomes one
group:

```{r composite}
constraint <- derive_split_constraints(
  g, mode = "composite", strategy = "strict",
  via = c("Subject", "Site", "Platform")
)
grouping_vector(constraint)
```

# Pairwise relation: genetic relatedness

Some leakage is pairwise and continuous rather than a clean grouping. Genetic
relatedness is the canonical example: a kinship coefficient — typically from a
tool such as KING or PLINK — links *pairs* of subjects.
`relatedness_edges_from_kinship()` takes such a pair table, keeps pairs at or
above a threshold, and emits `subject_related_to` edges; `mode = "relatedness"`
then groups by transitive closure over those edges (so a chain of related
individuals lands in one group).

```{r relatedness}
# A kinship table over subject pairs (one sample per subject here for clarity).
# P1-P2 and P2-P3 clear the threshold and chain together; P5-P6 form a second
# related pair; P1-P4 is too weak to count.
kin <- data.frame(
  id1     = c("P1", "P2", "P1", "P5"),
  id2     = c("P2", "P3", "P4", "P6"),
  kinship = c(0.25, 0.20, 0.02, 0.30),
  stringsAsFactors = FALSE
)
rel_edges <- relatedness_edges_from_kinship(kin, threshold = 0.1)

meta_r <- data.frame(
  sample_id  = paste0("S", 1:6),
  subject_id = paste0("P", 1:6),
  stringsAsFactors = FALSE
)
samples  <- create_nodes(meta_r, "Sample", "sample_id")
subjects <- create_nodes(meta_r, "Subject", "subject_id")
belongs  <- create_edges(meta_r, "sample_id", "subject_id",
                         "Sample", "Subject", "sample_belongs_to_subject")

g_rel <- build_dependency_graph(list(samples, subjects), list(belongs, rel_edges))

rel_groups <- grouping_vector(derive_split_constraints(g_rel, mode = "relatedness"))
rel_groups
```

The grouping is a transitive closure over the `subject_related_to` edges. The
network below draws those edges between subjects, coloured by the relatedness
group each subject (and therefore its samples) lands in: the P1–P2–P3 chain
becomes one group even though P1 and P3 were never linked directly, P5–P6 form a
second, and the unrelated P4 stands alone.

```{r rel-plot, fig.width = 6.5, fig.height = 4.5}
subject_group <- setNames(rel_groups[meta_r$sample_id], meta_r$subject_id)
kept_pairs <- kin[kin$kinship >= 0.1, c("id1", "id2")]
rel_net <- igraph::graph_from_data_frame(
  kept_pairs, directed = FALSE,
  vertices = data.frame(name = meta_r$subject_id)
)

palette_rel <- c("#4C78A8", "#F58518", "#54A24B", "#B279A2")
set.seed(1)
plot(rel_net,
     vertex.color       = palette_rel[as.integer(factor(subject_group[igraph::V(rel_net)$name]))],
     vertex.size        = 34,
     vertex.label.color = "white",
     vertex.label.font  = 2,
     edge.color         = "grey60",
     edge.width         = 2,
     main               = "Relatedness clusters (kinship >= 0.1)")
```

The threshold is the key knob, and it belongs to the *edge-building* step, not
the grouping. Raising it drops weaker links: at `0.22` the P2–P3 pair (kinship
`0.20`) no longer qualifies, so that chain breaks and P3 splits into its own
group, while the stronger P5–P6 pair is untouched:

```{r rel-threshold}
rel_strict <- relatedness_edges_from_kinship(kin, threshold = 0.22)
g_rel_strict <- build_dependency_graph(list(samples, subjects), list(belongs, rel_strict))

grouping_vector(derive_split_constraints(g_rel_strict, mode = "relatedness"))
```

# Pairwise relation: spatial proximity

Spatial proximity works the same way over sample coordinates — for example spot
locations from spatial transcriptomics, positions on a tissue slide, or
geographic site coordinates. `spatial_edges_from_coords()` connects samples
within a radius (Euclidean distance over the coordinate columns), and
`mode = "spatial"` groups the resulting connected components.

```{r spatial}
# Two spatial clusters. Cluster 1 (S1-S3) is a chain: neighbouring pairs are
# within the radius, but the endpoints are not.
coords <- data.frame(
  sample_id = paste0("S", 1:6),
  x = c(0, 1, 2,  6.0, 6.9, 6.2),
  y = c(0, 1, 0,  6.0, 6.6, 5.3),
  stringsAsFactors = FALSE
)
adj_edges <- spatial_edges_from_coords(coords, radius = 1.5)

meta_s <- data.frame(
  sample_id  = paste0("S", 1:6),
  subject_id = paste0("P", 1:6),
  stringsAsFactors = FALSE
)
samples_s  <- create_nodes(meta_s, "Sample", "sample_id")
subjects_s <- create_nodes(meta_s, "Subject", "subject_id")
belongs_s  <- create_edges(meta_s, "sample_id", "subject_id",
                           "Sample", "Subject", "sample_belongs_to_subject")

g_sp <- build_dependency_graph(list(samples_s, subjects_s), list(belongs_s, adj_edges))

sp_groups <- grouping_vector(derive_split_constraints(g_sp, mode = "spatial"))
sp_groups
```

Plotting the coordinates, drawing the within-radius adjacency edges in grey, and
colouring points by the derived group makes the transitive closure concrete:
S1–S2 and S2–S3 are each within the `1.5` radius, so all three share a group
even though S1 and S3 are `2` units apart and were never linked directly. Every
sample in the second cluster is likewise reachable from the others, while the
two clusters are far enough apart to stay separate:

```{r sp-plot, fig.width = 6.5, fig.height = 5}
sp_grp <- factor(sp_groups[coords$sample_id])
row_of <- setNames(seq_len(nrow(coords)), coords$sample_id)
from_i <- row_of[sub("^sample:", "", adj_edges$data$from)]
to_i   <- row_of[sub("^sample:", "", adj_edges$data$to)]
palette_sp <- c("#4C78A8", "#F58518")

plot(coords$x, coords$y, type = "n", asp = 1, xlab = "x", ylab = "y",
     main = "Spatial groups (radius = 1.5)")
segments(coords$x[from_i], coords$y[from_i],
         coords$x[to_i],   coords$y[to_i], col = "grey60", lwd = 2)
points(coords$x, coords$y, pch = 19, cex = 3.5, col = palette_sp[as.integer(sp_grp)])
text(coords$x, coords$y, labels = coords$sample_id, col = "white", cex = 0.8, font = 2)
legend("topleft", legend = levels(sp_grp), pch = 19,
       col = palette_sp[seq_along(levels(sp_grp))], title = "Spatial group", bty = "n")
```

# Deriving on a subset is leakage-safe

Real splits are derived on a *subset* of samples — the training rows, say. For
pairwise (and composite) modes this raises a subtle question: if a sample that
*bridges* two others is left out of the subset, could those two still inherit a
shared group from the full graph? They do not. When you pass `samples =`,
grouping is recomputed within that subset, so structure that exists only through
an excluded sample never leaks across the split.

The spatial chain makes this visible. S1 and S3 shared a group only because S2
bridged them; ask for S1 and S3 alone, and they correctly fall into separate
groups:

```{r subset-scoping}
grouping_vector(
  derive_split_constraints(g_sp, mode = "spatial", samples = c("S1", "S3"))
)
```

A subset is sometimes better expressed as a graph in its own right, for example
when you want to validate it, plot it, or hand it around. `subset_graph()` does
that, and it uses the same rule, so the grouping agrees with `samples =`:

```{r subset-graph}
g_two <- subset_graph(g_sp, samples = c("S1", "S3"), graph_name = "S1 + S3")
grouping_vector(derive_split_constraints(g_two, mode = "spatial"))
```

# Combining a pairwise relation with a direct one

Pairwise relations are not a separate world: either of them can be listed in a
composite `via` alongside the direct relations, and all of them feed the same
connected-component search. That matters when two different mechanisms each link
a different pair of samples, and severing only one would leave the other open.

Here relatedness links two subjects that share no batch, while batch links two
samples from unrelated subjects. Neither relation alone separates the cohort
correctly; the composite does:

```{r composite-pairwise}
meta_c <- data.frame(
  sample_id  = paste0("S", 1:4),
  subject_id = paste0("P", 1:4),
  batch_id   = c("B1", "B1", "B2", "B3"),
  stringsAsFactors = FALSE
)
kin <- data.frame(id1 = "P2", id2 = "P3", kinship = 0.25, stringsAsFactors = FALSE)

g_mixed <- build_dependency_graph(
  list(create_nodes(meta_c, "Sample", "sample_id"),
       create_nodes(meta_c, "Subject", "subject_id"),
       create_nodes(meta_c, "Batch", "batch_id")),
  list(create_edges(meta_c, "sample_id", "subject_id",
                    "Sample", "Subject", "sample_belongs_to_subject"),
       create_edges(meta_c, "sample_id", "batch_id",
                    "Sample", "Batch", "sample_processed_in_batch"),
       relatedness_edges_from_kinship(kin, threshold = 0.1))
)

# batch alone: {S1,S2} {S3} {S4}   relatedness alone: {S2,S3}
grouping_vector(derive_split_constraints(g_mixed, "batch"))
grouping_vector(derive_split_constraints(g_mixed, "relatedness"))

# combined, the chain S1-S2 (batch) - S3 (kinship) becomes one group:
mixed <- derive_split_constraints(g_mixed, "composite", via = c("batch", "relatedness"))
grouping_vector(mixed)
mixed$metadata$via
```

Under `strategy = "rule_based"` a pairwise source behaves as a fallback instead:
a sample takes the first relation in `priority` that gives it a non-singleton
group, so relations never chain.

Whichever route you take, the threshold that produced the edges is recorded on
the graph, per relation, and is serialised with it — so a reader of the file can
see which cut created the grouping:

```{r threshold-provenance}
g_mixed$metadata$edge_sources$subject_related_to[c("threshold", "metric")]
```

A spec derived from a *single* pairwise mode also carries that threshold as a
scalar in its own metadata. A composite spec does not, because several
relations (each with its own cut) may have contributed; the graph's
`edge_sources` is the complete record in that case:

```{r threshold-in-spec}
as_split_spec(derive_split_constraints(g_mixed, "relatedness"),
              graph = g_mixed)$metadata[c("threshold", "threshold_metric")]
```

# Thresholds are inputs, not modeling

Because the threshold (kinship cutoff, spatial radius) is applied up front in the
edge-building helpers, it is a *derivation input*, not a modeling choice:
`splitGraph` forms groups over whatever edges survive and never computes folds
itself. The resulting `split_spec` is handed to a downstream consumer for
execution, exactly as with every other mode — see the *adapter-cookbook* and
*cross-language-handoff* vignettes for that step.
