---
title: "Modeling site, platform, relatedness, and spatial structure"
author: "Selçuk Korkmaz"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Modeling site, platform, relatedness, and spatial structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(splitGraph)
```

Beyond the classic subject / batch / study / time relations, `splitGraph`
models four further sources of leakage: multi-**site** studies, sequencing /
measurement **platform**, and two genuinely *pairwise* relations — genetic
**relatedness** and **spatial** proximity. This vignette shows how to build and
group by each.

# Cluster-style relations: site, platform, assay

`graph_from_metadata()` auto-detects `site_id`, `platform_id`, and `assay_id`
columns and builds the corresponding typed nodes and edges. Each then has its
own constraint mode.

```{r cluster}
meta <- data.frame(
  sample_id   = paste0("S", 1:6),
  subject_id  = c("P1", "P1", "P2", "P2", "P3", "P3"),
  site_id     = c("NYC", "NYC", "BOS", "BOS", "NYC", "BOS"),
  platform_id = c("illumina", "illumina", "nanopore", "nanopore", "illumina", "nanopore"),
  assay_id    = c("rnaseq", "rnaseq", "rnaseq", "wgs", "wgs", "wgs"),
  stringsAsFactors = FALSE
)

g <- graph_from_metadata(meta, graph_name = "structure-demo")

grouping_vector(derive_split_constraints(g, mode = "site"))
grouping_vector(derive_split_constraints(g, mode = "platform"))
grouping_vector(derive_split_constraints(g, mode = "assay"))
```

Each of these annotations is also carried into `split_spec` as a blocking
variable, and any of them can participate in a composite derivation:

```{r composite}
constraint <- derive_split_constraints(
  g, mode = "composite", strategy = "strict",
  via = c("Subject", "Site", "Platform")
)
grouping_vector(constraint)
```

# Pairwise relation: genetic relatedness

Some leakage is pairwise and continuous rather than a clean grouping. Genetic
relatedness is the canonical example: a kinship coefficient links *pairs* of
subjects. `relatedness_edges_from_kinship()` keeps pairs at or above a threshold
and emits `subject_related_to` edges; `mode = "relatedness"` then groups by
transitive closure over those edges (so a chain of related individuals lands in
one group).

```{r relatedness}
# P1-P2 and P2-P3 are related above threshold; P4 is unrelated.
kin <- data.frame(
  id1 = c("P1", "P2", "P1"),
  id2 = c("P2", "P3", "P4"),
  kinship = c(0.25, 0.20, 0.02),
  stringsAsFactors = FALSE
)
rel_edges <- relatedness_edges_from_kinship(kin, threshold = 0.1)

meta_r <- data.frame(
  sample_id  = paste0("S", 1:4),
  subject_id = c("P1", "P2", "P3", "P4"),
  stringsAsFactors = FALSE
)
samples  <- create_nodes(meta_r, "Sample", "sample_id")
subjects <- create_nodes(meta_r, "Subject", "subject_id")
belongs  <- create_edges(meta_r, "sample_id", "subject_id",
                         "Sample", "Subject", "sample_belongs_to_subject")

g_rel <- build_dependency_graph(list(samples, subjects), list(belongs, rel_edges))

# S1, S2, S3 (P1-P2-P3 related) group together; S4 (P4) is separate.
grouping_vector(derive_split_constraints(g_rel, mode = "relatedness"))
```

# Pairwise relation: spatial proximity

Spatial proximity works the same way over sample coordinates.
`spatial_edges_from_coords()` connects samples within a radius (Euclidean
distance over the coordinate columns), and `mode = "spatial"` groups the
resulting connected components.

```{r spatial}
coords <- data.frame(
  sample_id = paste0("S", 1:4),
  x = c(0, 1, 2, 40),
  y = c(0, 0, 0, 40),
  stringsAsFactors = FALSE
)
adj_edges <- spatial_edges_from_coords(coords, radius = 1.5)

meta_s <- data.frame(
  sample_id  = paste0("S", 1:4),
  subject_id = paste0("P", 1:4),
  stringsAsFactors = FALSE
)
samples_s  <- create_nodes(meta_s, "Sample", "sample_id")
subjects_s <- create_nodes(meta_s, "Subject", "subject_id")
belongs_s  <- create_edges(meta_s, "sample_id", "subject_id",
                           "Sample", "Subject", "sample_belongs_to_subject")

g_sp <- build_dependency_graph(list(samples_s, subjects_s), list(belongs_s, adj_edges))

# S1-S2-S3 form a proximity chain; S4 is isolated.
grouping_vector(derive_split_constraints(g_sp, mode = "spatial"))
```

Because the thresholds are applied up front in the edge-building helpers, they
are *derivation inputs*, not modeling choices: `splitGraph` forms groups over
whatever edges survive, and never computes folds itself. The resulting
`split_spec` is handed to a downstream consumer for execution, exactly as with
every other mode.
