---
title: "Quick start: from a metadata table to a split_spec in ten minutes"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Quick start: from a metadata table to a split_spec in ten minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

splitGraph turns a sample-level metadata table into a **typed dependency
graph**, validates it, derives a **split constraint** (which samples must stay
on the same side of any train/test split), and emits a tool-agnostic
**`split_spec`** that downstream resampling tools consume. It never creates
folds itself. This vignette is the shortest complete path; the other vignettes
go deeper.

```{r load}
library(splitGraph)
```

## 1. A metadata table

One row per sample. Any of the canonical columns below may be present; absent
ones are skipped.

```{r meta}
meta <- data.frame(
  sample_id    = paste0("S", 1:8),
  subject_id   = c("P1", "P1", "P2", "P2", "P3", "P3", "P4", "P4"),
  batch_id     = c("B1", "B1", "B1", "B2", "B2", "B2", "B3", "B3"),
  site_id      = c("NYC", "NYC", "NYC", "BOS", "BOS", "BOS", "BOS", "BOS"),
  timepoint_id = rep(c("T0", "T1"), 4),
  time_index   = rep(c(0, 1), 4),
  outcome_id   = c("case", "case", "ctrl", "ctrl", "case", "case", "ctrl", "ctrl"),
  stringsAsFactors = FALSE
)
```

| Canonical column | Creates | Used by mode |
|---|---|---|
| `sample_id` (required) | `Sample` nodes | all |
| `subject_id` | `Subject` + `sample_belongs_to_subject` | `subject`, `relatedness` |
| `batch_id` | `Batch` + `sample_processed_in_batch` | `batch` |
| `study_id` | `Study` + `sample_from_study` | `study` |
| `site_id` | `Site` + `sample_collected_at_site` | `site` |
| `region_id` | `Region` + `sample_located_in_region` | `region` |
| `platform_id` | `Platform` + `sample_run_on_platform` | `platform` |
| `assay_id` | `Assay` + `sample_measured_by_assay` | `assay` |
| `timepoint_id` (+ `time_index`) | `Timepoint` + `sample_collected_at_timepoint` (+ `timepoint_precedes`) | `time` |
| `outcome_id` / `outcome_value` | `Outcome` + `sample_has_outcome` | `stratum` annotation |

Identifier columns may be character, factor, or numeric. Use `columns =` to map
your own column names onto the canonical ones.

## 2. Build and validate

```{r build}
g <- graph_from_metadata(meta, graph_name = "quick-start")
g
validate_graph(g)
```

The report lists structural errors (which stop the build), semantic warnings,
and leakage advisories such as *repeated subject samples*. Nothing here decides
a split yet; it tells you what structure exists.

## 3. Choose a constraint mode

| Your question | `mode` | What ends up in one group |
|---|---|---|
| Same individual measured several times? | `"subject"` | all samples of a subject |
| Processing batch / plate / run effects? | `"batch"` | all samples of a batch |
| Several studies or cohorts pooled? | `"study"` | all samples of a study |
| Multi-centre collection? | `"site"` | all samples of a site |
| Tissue region, sequencing platform, assay? | `"region"`, `"platform"`, `"assay"` | likewise |
| Longitudinal, want train-before-test ordering? | `"time"` | samples of a timepoint, plus `order_rank` |
| Genetic relatives or spatial neighbours (continuous similarity)? | `"relatedness"`, `"spatial"` | connected components over thresholded edges (see `?pairwise_edges`) |
| Several of the above at once? | `"composite"` | strict: one group per connected component over all chosen relations; rule-based: first available relation in a priority order |

```{r derive}
constraint <- derive_split_constraints(g, mode = "composite", via = c("subject", "batch"))
constraint
head(as.data.frame(constraint), 4)
```

`grouping_vector()` gives the group per sample as a named vector, the handle
most tools want:

```{r grouping}
grouping_vector(constraint)
```

## 4. Emit the split_spec

```{r spec}
spec <- as_split_spec(constraint, graph = g)
spec
validate_split_spec(spec)
```

Passing `graph = g` enriches the spec with everything the constraint did not
use as the primary grouping: blocking columns (`batch_group`, `site_group`,
...), the ordering (`order_rank`) and the `stratum` annotation from the
`Outcome` nodes, so a downstream tool can block, order, or stratify without
touching the graph.

```{r spec-columns}
names(spec$sample_data)
spec$block_vars
spec$time_var
spec$stratum_var
```

## 5. Hand off

Write the spec to JSON for another session, another language, or an archive:

```{r write, eval = requireNamespace("jsonlite", quietly = TRUE)}
path <- tempfile(fileext = ".json")
write_split_spec(spec, path)
validate_split_spec_json(path)
```

Then consume it. In R with bioLeak:

```{r bioleak, eval = FALSE}
bioLeak::as_leaksplits(spec, data = my_frame, outcome = "y")
```

In Python (the reader ships with the package and needs only the standard
library):

```python
from splitspec import load_split_spec
spec = load_split_spec("split_spec.json")
spec.groups()      # for GroupKFold(groups=...)
spec.strata()      # for StratifiedGroupKFold(y=...)
spec.order_ranks() # sort before TimeSeriesSplit
```

See `vignette("cross-language-handoff")` for the complete Python path,
`vignette("adapter-cookbook")` for rsample and base-R adapters, and
`vignette("faq-design-notes")` for the reasoning behind the design.

```{r cleanup, include = FALSE}
if (exists("path")) unlink(path)
```
